Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Updates

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) rather than a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative 
	structure learning approach that combines domain knowledge with data-driven
	insights. Our method leverages Conditional Independence (CI) testing to
	iteratively identify variable pairs where an edge is either missing or
	superfluous. Based on this information, we can choose to add missing
	edges with appropriate orientation based on domain knowledge or remove
	unnecessary ones. We also give a method to rank potential new edges
	based on their impact on the overall model fit. Empirical results show
	that this iterative approach achieves performance comparable to
	automated algorithms if the orientation of new edges is correctly
	determined in at least one out of two cases. Additionally, we show
	that, in the absence of an expert, a Large Language Model (LLM) can be
	used to determine edge orientations. We provide an implementation of
	our approach as both a webtool: <redacted for review> and a Python
	package <redacted for review>.

@@

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships
among random variables using observational data.

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, such as constraint-based methods like PC
algorithm \citep{Spirtes2001} and Fast Causal Inference \citep{Spirtes2000}),
score-based methods such as Hill-Climb Search and Greedy Equivalence Search
\citep{Chickering2002}, and continuous optimization-based methods like NO TEARS
\citep{Zheng2018} and DAGMA \citep{Bello2022}. Despite significant progress in
automated causal discovery, their adoption in applied research has been
limited. Researchers often prefer to manually construct DAGs based on their
domain expertise \citep{Tennant2020, Petersen2021}. We attribute this
preference to several challenges with existing causal discovery algorithms in
practical settings:

@@

Figure~\ref{fig:intro} highlights some of these issues. On the other hand,
constructing DAG based solely on domain expertise can be error prone as it
requires us to distinguish between different causal structures, such as direct
or indirect effects or causal mediation. Given these complexities, it is
important to validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the CI statements implied by the
DAG hold in the data \citep{Ankan2021}. Specifically, each missing edge between
a pair of variables in the DAG implies a CI statement, which can be tested in
the dataset using a statistical test. If we find any violation to a CI
statement, we can fix it by adding an edge between the variables in the DAG. We
also need to decide the orientation of this edge based on domain expertise.

@@

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during the DAG construction process,
rather than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iterations, we use CI testing to identify pairs of variables
where an edge is either missing or superfluous. These pairs are then presented
to the expert, who determines whether to add or remove these edges. The
orientation of the edge when adding is determined based on the domain
expertise. As at any iteration, multiple changes to the DAG are possible. We
also give a way to rank potential new edges additions to the DAG using effect
size measures for CI tests to address the most severe violations of CI tests
first. As a result this approach acts like a guide to the expert when manually
constructing their model while ensuring that the model is consistent with the
data.

% When manually constructing models, it is important to test whether the
% resulting DAG accurately represents the dataset. Based on this evaluation and
% domain knowledge, we can make further modification to the model. One common
% approach for assessing DAGs is to check whether the Conditional Independences
% (CIs) implied by the DAG hold in the data \citep{Ankan2021}. If we find
% violations to these CIs, we can use our domain knowledge to make appropriate
% modifications to the DAG. However, this CI testing approach has a few
% limitations:
% 
% \begin{enumerate}
% \item As CIs are only implied by missing edges in the model, this approach does 
% 	not test whether the existing edges are correct.
% \item Determining whether a CI holds in data is based on a p-value threshold. This can 
% 	be unreliable, for example, we get a significant p-value for even very weak 
% 	relationship if the sample size is high.
% \item The number of implied CIs can be large making it difficult to manually
% 	check all of them. For example, a moderately sized alarm network
% 	\citep{Beinlich1989} with $ 37 $ nodes, implies $ 287 $ CIs.
% \end{enumerate}
% 
% To tackle these limitations, we draw inspiration from modification indices in
% specification search. Modification indices provide ranking for potential model
% modification based on their impact on the model's fit, allowing us to
% prioritize most meaningful modifications.  Similarly, we propose a method for
% ranking CI test violations to help us focus on the most critical
% inconsistencies in the model. To determine this ranking we utilize a measure of
% conditional association between variables in combination with p-values from the
% CI test. This approach addresses all three of the issues outlined above. The
% measure of association allows us to evaluate the validity of existing edges. It
% reduces our reliance on p-values alone for deciding whether the CI test holds.
% The ranking helps us prioritize the most critical violations, making it feasible 
% to focus on the areas that most improve the model's fit.

@@

>>>>

	Causal discovery has received significant attention in the Directed
	Acyclic Graphs (DAGs) literature, leading to the development of
	numerous automated algorithms for learning DAGs from data. However,
	their adoption in applied domains remain limited, as researchers often
	prefer to construct DAGs manually based on domain knowledge. This
	preference arises due to several practical challenges with automated
	algorithms, such as their tendency to make obvious errors and output
	Markov Equivalence Classes (MECs) instead of a DAG. To assist
	researchers in constructing DAGs manually, we propose an iterative
	structure learning approach that combines domain knowledge with
	data-driven insights. Our method leverages Conditional Independence
	(CI) testing to iteratively identify variable pairs where an edge is
	either missing or superfluous. Based on this information, we can choose
	to add missing edges with appropriate orientation based on domain
	knowledge or remove unnecessary ones. We also give a method to rank
	these missing edges based on their impact on the overall model fit.
	Empirical results show that this iterative approach achieves
	performance comparable to automated algorithms if the orientation of
	new edges is correctly determined in at least one out of two cases.
	Additionally, we show that, in the absence of an expert, a Large
	Language Model (LLM) can be used to determine edge orientations. We
	provide an implementation of our approach as both a web tool: <redacted
	for review> and a Python package <redacted for review>.

@@

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. Causal discovery methods aim to discover such relationships among
random variables using observational data. In the DAG literature, the primary
focus has been on developing automated algorithms to learn causal structures
from datasets. These efforts have led to numerous causal discovery algorithms,
such as constraint-based methods like PC algorithm \citep{Spirtes2001} and Fast
Causal Inference \citep{Spirtes2000}), score-based methods such as Hill-Climb
Search and Greedy Equivalence Search \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. Despite significant progress in automated causal discovery,
their adoption in applied research has been limited. Researchers often prefer
to manually construct DAGs based on their domain expertise \citep{Tennant2020,
Petersen2021}. We attribute this preference to several challenges with existing
causal discovery algorithms in practical settings:

@@

Figure~\ref{fig:intro} highlights some of these issues. On the other hand,
constructing DAG based solely on domain expertise can be error prone as it
requires us to distinguish between different causal structures, such as direct
or indirect effects or causal mediation. Given these complexities, it is
important to validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the CI statements implied by the
DAG hold in the data \citep{Ankan2021}. Specifically, each missing edge between
a pair of variables in the DAG implies a CI statement, which can be tested in
the dataset using a statistical test. If we find any violation to CI
statements, we can make modifications to our model accordingly to fix the
violations.

@@

In this paper, we propose a structure learning method that leverages this CI
testing approach to assist researchers during manual DAG construction, rather
than just validating the model. Our approach follows an iterative
model-building strategy that integrates domain expertise with CI testing on
data. At each iteration, we use CI testing to identify pairs of variables where
an edge is either missing or superfluous. We can then use this information
along with domain expertise to determines whether to add or remove these edges.
The orientation of the edge when adding is determined based on the domain
expertise. As at any iteration, multiple changes to the DAG are possible. We
also give a way to rank potential new edges additions to the DAG using effect
size measures for CI tests to address the most severe violations of CI tests
first. As a result this approach acts like a guide to the expert when manually
constructing their model while ensuring that the model is consistent with the
data.

@@

\subsection{Model Testing Using CI Tests}
